\section{Owlready2}

\subsection{What is Owlready2?}

Owlready2 is a Python package for \textbf{ontology-oriented programming}: load,
create, modify, and reason over OWL~2 ontologies as ordinary Python objects
(\href{https://owlready2.readthedocs.io/en/latest/intro.html}{Owlready2 documentation}).
It fits our stack (\texttt{Python 3.11+}, \texttt{uv}) better than a separate
Java GUI editor.

Core capabilities:
\begin{itemize}
  \item Import/export OWL in RDF/XML, OWL/XML, N-Triples
  \item Define classes, properties, and individuals in Python
  \item Run \textbf{HermiT} or \textbf{Pellet} via \texttt{sync\_reasoner()}
  \item SPARQL queries over an embedded RDF quadstore (SQLite)
  \item Optional \textbf{PyMedTermino2} submodule for medical terminologies (UMLS)
\end{itemize}

\subsection{Python constructs and OWL}

\begin{table}[htbp]
  \centering
  \small
  \caption{Owlready2 API and OWL equivalents}
  \label{tab:owlready2-owl}
  \renewcommand{\arraystretch}{1.2}
  \begin{tabularx}{\textwidth}{@{}l X@{}}
    \toprule
    \textbf{Python (Owlready2)} & \textbf{OWL concept} \\
    \midrule
    \texttt{class Diagnosis(Thing): ...} & \texttt{owl:Class} \\
    \texttt{class worksFor(ObjectProperty): ...} & \texttt{owl:ObjectProperty} \\
    \texttt{domain = [Person]} / \texttt{range = [Organization]} &
    \texttt{rdfs:domain}, \texttt{rdfs:range} \\
    \texttt{Diagnosis("mention\_1")} & Named individual (ABox) \\
    \texttt{mention.worksFor = [org]} & Object property assertion \\
    \texttt{mention.label = ["trào ngược..."]} & \texttt{rdfs:label} \\
    \texttt{sync\_reasoner()} & HermiT classification + inferred types \\
    \texttt{onto.save()} & Export RDF/XML or Turtle \\
    \bottomrule
  \end{tabularx}
\end{table}

\subsection{Minimal example}

\begin{verbatim}
from owlready2 import *

onto = get_ontology("http://onto-med-rag.example/clinical.owl")

with onto:
    class Diagnosis(Thing): pass
    class Drug(Thing): pass
    class mapsToICD(ObjectProperty):
        domain = [Diagnosis]
        range  = [Thing]

    class Mention(Diagnosis):
        pass

mention = Mention("trào_ngược")
mention.label = ["bệnh trào ngược dạ dày - thực quản"]

with onto:
    sync_reasoner()   # requires Java (JVM) for HermiT

onto.save("data/kb/clinical_overlay.owl")
\end{verbatim}

After \texttt{sync\_reasoner()}, Owlready2 updates instance and class hierarchies
from inferred facts---the same deterministic reasoning described in
Table~\ref{tab:layer-cake}, but callable from pipeline code
(\href{https://owlready2.readthedocs.io/en/latest/reasoning.html}{Reasoning docs}).

\subsection{Reasoning workflow}

\begin{enumerate}
  \item Define classes, properties, restrictions, and disjointness in Python.
  \item Create individuals and assert facts (or load an existing OWL file with
        \texttt{onto.load()}).
  \item Call \texttt{sync\_reasoner()} (HermiT, default) or
        \texttt{sync\_reasoner\_pellet()}.
  \item Inspect reclassified instances (\texttt{instance.\_\_class\_\_}) and inferred
        superclasses; fix inconsistent classes before production use.
  \item Save or query via SPARQL for \texttt{src/linking/} and
        \texttt{src/ranking/}.
\end{enumerate}

HermiT and Pellet run on the JVM---install Java before reasoning
(\href{https://owlready2.readthedocs.io/en/latest/reasoning.html}{Owlready2 reasoning}).

\subsection{Loading external ontologies}

ICD-10 and RxNorm can be loaded when available as OWL/RDF:

\begin{verbatim}
from owlready2 import *

onto = get_ontology("data/kb/clinical_overlay.owl").load()
# or merge imports from a standard terminology file
icd = get_ontology("data/kb/icd10.owl").load()
\end{verbatim}

Owlready2 keeps a quadstore in memory or on disk; lists on ontology objects are
modified in place and the store updates automatically
(\href{https://owlready2.readthedocs.io/en/latest/intro.html}{Introduction}).

\subsection{Why Owlready2 for \texttt{onto-med-rag}}

\begin{itemize}
  \item Same language as the NLP pipeline---no context switch to Java/GUI.
  \item Scriptable: build KB overlays, run reasoner checks in CI, export Turtle
        from \texttt{uv run} scripts.
  \item Can attach Python methods to OWL classes (e.g.\ validation helpers for
        assertion flags).
  \item PyMedTermino2 integration for UMLS/medical codes when we extend beyond
        competition ICD/RxNorm files
        (\href{https://pypi.org/project/owlready2/}{PyPI}).
\end{itemize}

\subsection{Next step for our project}

See Section~5 for how Owlready2 connects to the competition pipeline, JSON output
schema, and phased roadmap.
